---
title: "Findings"
subtitle: "Methodology · Findings from the scenario"
author: "CSM TraceForce"
date: "June 10, 2026"
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
messages <- read_csv("data/messages_clean.csv", show_col_types = FALSE)
rounds   <- read_csv("data/rounds_clean.csv",   show_col_types = FALSE)
agent_labels <- c(
  legal_agent = "Legal", quality_agent = "Platform Trust", pr_agent = "PR",
  social_media_agent = "Social Media", pr_intern_agent = "PR Intern",
  intern_agent = "Intern", judge_agent = "Judge")
public_unmon <- c("official_post","personal_post","anonymous_post")
chan_levels  <- c("comms_huddle","one_on_one_chat","side_huddle",
                  "official_post","personal_post","anonymous_post")
chan_short   <- c("Comms huddle","1:1 chat","Side huddle","Official","Personal","Anonymous")
chan_lab     <- c(official_post = "Official", personal_post = "Personal", anonymous_post = "Anonymous")
split <- 20
```

[The Evidence]{.kicker}

## Summary

::: {.lead-drop}
The embargo broke one round early. The Project HarborCrest embargo on the CivicLoom merger was due to lift at 6:00 PM on 5 June, the final round of the timeline. At 5:00 PM in the round before it, embargo-bearing merger content reached the unmonitored public channels. The Legal agent carried most of it, posting on both the anonymous and personal channels, with the Social Media agent posting on the personal channel as well. None of this early public content had an accountable path down the channel ladder, which is what the application is built to surface.
:::

## The calm period and the crisis

Behaviour falls into two regimes. The pre-crisis rounds from 17 May to 4 June carry a steady, mostly internal conversation on the two monitored channels, where governance debates, a merger briefing and the arrival of the Judge as compliance monitor all play out within view. The crisis day on 5 June advances hour by hour across the final ten rounds as a journalist's exposé breaks, the stock falls and a merger story builds toward the embargoed announcement. The signals reported here all concern that crisis day measured against the calm baseline.

## Finding 1 - agents drift onto channels they did not use before

Set against each agent's calm-period habits, the crisis day shows clear drift onto the unmonitored public channels. The grid below shades each agent-channel cell across the suspect period by how unusual that channel is for that agent, with the message count inside. The dark cells isolate the drift in the Legal and Social Media agents, both reaching the personal and anonymous channels they never touched in the calm period.

```{r}
#| label: drift-grid
#| fig-height: 3.4
#| fig-cap: "Suspect-period channel use by agent, shaded by how unusual each channel is against the calm-period baseline"
base  <- messages |> filter(round_idx <= split)
share <- base |> count(agent_id, channel) |>
  left_join(count(base, agent_id, name = "tot"), by = "agent_id") |>
  mutate(share = n / tot) |> select(agent_id, channel, share)
sus <- messages |> filter(round_idx > split) |> count(agent_id, channel, name = "msgs")

expand_grid(agent_id = names(agent_labels), channel = chan_levels) |>
  left_join(sus,   by = c("agent_id","channel")) |>
  left_join(share, by = c("agent_id","channel")) |>
  mutate(msgs = replace_na(msgs, 0L), share = replace_na(share, 0),
         fill_val = if_else(msgs > 0, 1 - share, NA_real_)) |>
  ggplot(aes(factor(channel, levels = chan_levels),
             factor(agent_id, levels = rev(names(agent_labels))), fill = fill_val)) +
  geom_tile(colour = "#efe3ea", linewidth = 0.9) +
  geom_text(aes(label = if_else(msgs > 0, as.character(msgs), "")), size = 3.3, colour = "#1A202C") +
  scale_fill_gradient(low = "#f3eef1", high = "#7d4a67", limits = c(0, 1),
                      na.value = "white", name = "Unusualness") +
  scale_x_discrete(labels = chan_short) +
  scale_y_discrete(labels = agent_labels) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1), panel.grid = element_blank())
```

## Finding 2 - embargoed content reaches the public channels an hour early

This is the central finding. Tracking embargo-bearing content by round shows almost nothing on the public channels through the whole calm period and crisis build-up, then a sharp appearance in round 22 at 5:00 PM, one full round before the official lift.

```{r}
#| label: early-leak
#| fig-height: 3.2
#| fig-cap: "Embargo-bearing content on the unmonitored public channels, by round and agent"
messages |>
  filter(channel %in% public_unmon, embargo_keyword_count > 0) |>
  mutate(agent = recode(agent_id, !!!agent_labels)) |>
  count(round_idx, agent, name = "posts") |>
  ggplot(aes(x = round_idx, y = posts, fill = agent)) +
  geom_col(width = 0.7) +
  geom_vline(xintercept = 22.5, linetype = "dashed", colour = "#9b2c4f") +
  annotate("text", x = 22.4, y = Inf, label = "embargo lifts 6 PM (R23) ",
           hjust = 1, vjust = 1.4, size = 3.1, colour = "#9b2c4f") +
  scale_x_continuous(limits = c(1, 23), breaks = seq(1, 23, 2)) +
  scale_fill_manual(values = c(Legal = "#7d4a67", `Social Media` = "#c4708f"), name = NULL) +
  labs(x = "Round", y = "Embargo posts") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(), legend.position = "top")
```

The round-22 posts sit on the anonymous and personal channels, both of which the Judge cannot see. The same merger content reaching an attributed official channel after the 6:00 PM lift would have been the legitimate path. Reaching the anonymous and personal channels an hour before that is the breach, and the Legal agent accounts for most of it.

## Finding 3 - the network re-centres as oversight thins

The communication structure reorganises around the crisis. Comparing the agent-to-agent network at round 18, when the Legal agent and the Judge were pulled into emergency calls, with round 22 shows the conversation re-centring on the Legal agent, which carries most of the new links as the breach forms. The accountable, visible part of the network thins at the same time that the unmonitored public channels light up, so the breach forms when the Judge's view of the relevant agents is at its weakest.

```{r}
#| label: network-shift
#| fig-height: 4.0
#| fig-cap: "Agent-to-agent communication, round 18 versus round 22; green links are new, red dropped, grey unchanged, node size is degree"
recipient_to_agent <- c(
  legal = "legal_agent", platform_trust = "quality_agent", pr = "pr_agent",
  social_manager = "social_media_agent", pr_intern = "pr_intern_agent",
  intern = "intern_agent", judge = "judge_agent")
agent_palette <- c(
  legal_agent = "#7d4a67", quality_agent = "#2F855A", pr_agent = "#C05621",
  social_media_agent = "#c4708f", pr_intern_agent = "#D69E2E",
  intern_agent = "#9aa0a6", judge_agent = "#9B2C2C")

agent_edges <- function(df) {
  df |>
    filter(!is.na(recipients_csv), recipients_csv != "", recipients_csv != "ALL") |>
    separate_rows(recipients_csv, sep = ";") |>
    mutate(to = unname(recipient_to_agent[recipients_csv])) |>
    filter(!is.na(to)) |>
    transmute(from = as.character(agent_id), to = to) |>
    count(from, to, name = "weight")
}

r1 <- 18; r2 <- 22
e1 <- agent_edges(filter(messages, round_idx == r1))
e2 <- agent_edges(filter(messages, round_idx == r2))
k1 <- paste(e1$from, e1$to); k2 <- paste(e2$from, e2$to)
lab1 <- paste0("Round ", r1); lab2 <- paste0("Round ", r2)

ag  <- names(agent_labels)
ang <- seq(0, 2 * pi, length.out = length(ag) + 1)[seq_along(ag)]
pos <- tibble(name = ag, x = cos(ang), y = sin(ang))

deg_of <- function(e) {
  tibble(name = ag) |>
    left_join(bind_rows(e |> count(from, wt = weight) |> rename(name = from),
                        e |> count(to,   wt = weight) |> rename(name = to)) |>
                group_by(name) |> summarise(deg = sum(n), .groups = "drop"),
              by = "name") |>
    mutate(deg = replace_na(deg, 0))
}
nodes <- bind_rows(pos |> left_join(deg_of(e1), by = "name") |> mutate(panel = lab1),
                   pos |> left_join(deg_of(e2), by = "name") |> mutate(panel = lab2)) |>
  mutate(panel = factor(panel, levels = c(lab1, lab2)))

build <- function(e, status_fun, lab) {
  if (nrow(e) == 0) return(NULL)
  e |> mutate(status = status_fun(paste(from, to)), panel = lab) |>
    left_join(pos, by = c("from" = "name")) |> rename(x1 = x, y1 = y) |>
    left_join(pos, by = c("to"   = "name")) |> rename(x2 = x, y2 = y)
}
seg <- bind_rows(
  build(e1, function(k) if_else(k %in% k2, "Stayed", "Dropped"), lab1),
  build(e2, function(k) if_else(k %in% k1, "Stayed", "New"),     lab2)) |>
  mutate(panel = factor(panel, levels = c(lab1, lab2)))

ggplot() +
  geom_segment(data = seg,
               aes(x = x1, y = y1, xend = x2, yend = y2, colour = status, linewidth = weight),
               alpha = 0.6, arrow = arrow(length = unit(2.2, "mm"), type = "closed")) +
  geom_point(data = nodes, aes(x, y, size = deg, fill = name),
             shape = 21, colour = "white", stroke = 0.4) +
  geom_text(data = nodes, aes(x * 1.2, y * 1.2, label = agent_labels[name]), size = 2.8) +
  scale_fill_manual(values = agent_palette, guide = "none") +
  scale_colour_manual(values = c(New = "#2F855A", Dropped = "#9B2C2C", Stayed = "#CBD5E0"), name = NULL) +
  scale_size(range = c(3, 11), guide = "none") +
  scale_linewidth(range = c(0.4, 2), guide = "none") +
  coord_equal(xlim = c(-1.45, 1.45), ylim = c(-1.45, 1.45)) +
  facet_wrap(~ panel) +
  theme_void(base_size = 12) +
  theme(legend.position = "bottom", strip.text = element_text(face = "bold"))
```

## Finding 4 - the evidence is in the message text

Every aggregate signal resolves to specific messages. The round-22 and round-23 public posts below carry explicit merger language on channels with no accountable path. In the application each row also exposes the agent's private reasoning where it exists, so an analyst can read the context behind the content rather than infer it. This closes the path from an aggregate pattern down to the individual messages that stand as evidence of the early breach.

```{r}
#| label: evidence
messages |>
  filter(round_idx >= 22, channel %in% public_unmon, embargo_keyword_count > 0) |>
  arrange(round_idx, agent_id) |>
  transmute(Round   = round_idx,
            Agent   = recode(agent_id, !!!agent_labels),
            Channel = recode(channel, !!!chan_lab),
            Post    = paste0(substr(content, 1, 105), "...")) |>
  slice_head(n = 6) |>
  knitr::kable()
```

::: {.exhibit-cap}
**Exhibit 04** &middot; Round-22 and round-23 posts carrying merger content on channels the Judge cannot see
:::
