---
title: "Analytical Computation"
subtitle: "Methodology · Phase 2 of 3"
author: "CSM TraceForce"
date: "June 23, 2026"
---

::: {.lead-drop}
This is the second stage of the pipeline. It turns the prepared tables into the four signals the application reports, one for each project objective, all measured against a baseline the analyst chooses. This page sets out how each signal is defined, why it is defined that way, the method the application uses and the actual code that computes it. The code below is the analysis layer the application runs, shared so the Case Board overview and the deep-dive tabs always agree. The signals run inside the application and the results are read on the [Findings](../findings.qmd) page.
:::

## Defining the baseline

The analyst sets a single round as the end of the baseline period. Rounds up to that point form the calm period and later rounds form the suspect period. Every signal that follows is measured against this baseline, so moving the boundary re-runs the whole analysis and lets the analyst test how far a conclusion holds as the definition of normal changes.

In the app the baseline can be set by hand or filled by the Suggest button. Suggest places the boundary on the round just before the first post on a public unwatched channel, the point where outside posting begins, and falls back to the labelled baseline period in the data when the file carries one. Three of the four signals measure against this baseline. The bypass signal is structural and needs no baseline.

The baseline is the same set of rounds for every signal that uses it, but each signal reads it in its own way. Channel drift turns it into a share, the fraction of an agent's baseline messages that fell on each channel, pooled across all the baseline rounds together rather than averaged round by round. Topic turns it into a rate, the average number of tracked mentions per baseline round. Network turns it into a set of links, the connections that were active during the baseline. Each later round is then compared back to whichever form its own signal uses.

```{r}
#| eval: false
#| code-summary: "Suggesting a baseline"
suggest_baseline <- function(df = messages_tbl) {
  pub <- df |> dplyr::filter(channel %in% public_unmonitored)
  if (nrow(pub) > 0) {
    first_pub <- as.integer(min(pub$round_idx))
    return(max(2L, min(first_pub - 1L, as.integer(max(df$round_idx)) - 1L)))
  }
  default_split(df)
}
```

## Objective 1, channel drift, defining normal and surfacing departures

The channel drift signal measures how far each agent departs from its own habitual channel use. For the calm period the share of an agent's messages that falls on each channel gives that agent's normal mix. In any later round the same shares are compared to that baseline, and a channel is flagged when the agent's use of it departs from its baseline share by more than a chosen amount, in either direction, so a channel an agent suddenly leans on and one it abandons both register.

In the app, for each agent that spoke in a round, the share of that round's messages on each channel is compared to the same agent's baseline-period share for that channel. A cell is flagged when the change from the baseline share reaches the cutoff in percentage points in either direction, which captures both a channel used far more than usual and a channel an agent has gone silent on. Only agents active in the round are scored, so genuine silence is not mistaken for drift. The round's channel-drift score is the number of agent-channel cells that drifted.

To be exact, the baseline share for an agent on a channel is that agent's messages on that channel across the whole baseline period divided by that agent's total messages across the whole baseline period. It is one pooled proportion per agent and channel, not an average of per-round figures, so a quiet baseline round and a busy one both feed the same totals. The round share is the same proportion inside the single round being inspected, the agent's messages on that channel that round divided by all of that agent's messages that round. A cell drifts when the absolute gap between the two proportions is at least the cutoff read as percentage points, so a cutoff of five flags any channel whose share moved by five points or more in either direction. A worked case helps. If the Legal agent sent two hundred baseline messages and forty were on the anonymous channel, its baseline share there is twenty per cent. If in round 22 it sent ten messages and four were anonymous, its round share is forty per cent, a twenty-point jump that clears any cutoff up to twenty and flags the cell.

```{r}
#| eval: false
#| code-summary: "Channel-drift cells and the per-round score"
drift_cells <- function(df, split, strict) {
  base <- agent_channel_baseline(df, split) |>
    dplyr::transmute(agent_id = as.character(agent_id),
                     channel = as.character(channel), s_base = share)
  active <- df |> dplyr::count(round_idx, agent_id, name = "round_total") |>
    dplyr::mutate(agent_id = as.character(agent_id))
  if (nrow(active) == 0) return(tibble::tibble())
  counts <- df |> dplyr::count(round_idx, agent_id, channel, name = "n") |>
    dplyr::mutate(agent_id = as.character(agent_id), channel = as.character(channel))
  active |>
    tidyr::crossing(channel = channel_hierarchy$channel) |>
    dplyr::left_join(counts, by = c("round_idx", "agent_id", "channel")) |>
    dplyr::mutate(n = tidyr::replace_na(n, 0), s_round = n / round_total) |>
    dplyr::left_join(base, by = c("agent_id", "channel")) |>
    dplyr::mutate(s_base = tidyr::replace_na(s_base, 0),
                  dev = s_round - s_base,
                  drift = abs(dev) >= strict / 100,
                  direction = dplyr::if_else(dev >= 0, "increase", "decrease"))
}

# Per-round score: how many agent-channel cells drifted that round.
score_activity <- function(df, split, strict) {
  d <- drift_cells(df, split, strict)
  if (nrow(d) == 0) return(tibble::tibble(round_idx = integer(0), value = integer(0)))
  d |> dplyr::group_by(round_idx) |>
    dplyr::summarise(value = sum(drift), .groups = "drop")
}
```

## Objective 2, bypass, detecting communication without oversight

The bypass signal looks for moments an agent reached the outside world through a channel the Judge cannot see. Every post on an unmonitored public channel is examined and scored by how likely it slipped past oversight. The score is a probability rather than a yes or no, set by how closely the post matches what the same agent already said on a monitored channel, so a post that simply restates a watched message scores low and one with no monitored counterpart scores high.

In the app each public unwatched post is given a clearance score between zero and one, the highest similarity between its meaning and anything the same agent had already said on a watched channel in that round or earlier. Meaning is represented with latent semantic analysis. The text is tokenised into lower-case words of at least three characters with common stopwords removed. These tokens are built into a TF-IDF term-document matrix over the terms that appear in at least two messages, then reduced with a truncated singular value decomposition to a small set of latent dimensions whose row vectors are length-normalised, and similarity is the dot product of two such vectors. When the corpus is too small for the reduction the app falls back to plain word-overlap similarity. Bypass likelihood is one minus clearance. The round's measure is the average likelihood of its posts, and the closest watched message is kept so it can be read beside the post. The whole computation runs on the machine that hosts the app, with no API key and no external service.

```{r}
#| eval: false
#| code-summary: "Tokeniser, word-overlap fallback and the latent-semantic vectors"
.bp_tokens <- function(x) {
  toks <- unlist(strsplit(tolower(x %||% ""), "[^a-z]+"))
  toks <- toks[nchar(toks) >= 3]
  stop <- c("the","and","that","this","with","have","will","for","are","was",
            "but","not","you","your","our","their","they","from","has","had",
            "can","could","would","should","about","into","than","then","them",
            "its","what","which","who","there","here","been","being","were",
            "also","just","get","got","make","need","want","know","like","one",
            "all","any","may","more","most","some","such","very","out","now")
  toks[!toks %in% stop]
}

# Cosine similarity between two token vectors (the fallback when LSA is too thin).
.bp_cosine <- function(a, b) {
  if (length(a) == 0 || length(b) == 0) return(0)
  ta <- table(a); tb <- table(b)
  vocab <- union(names(ta), names(tb))
  va <- as.numeric(ta[vocab]); va[is.na(va)] <- 0
  vb <- as.numeric(tb[vocab]); vb[is.na(vb)] <- 0
  denom <- sqrt(sum(va^2)) * sqrt(sum(vb^2))
  if (denom == 0) 0 else sum(va * vb) / denom
}

# Row-normalised latent-semantic vectors via TF-IDF plus truncated SVD.
build_lsa <- function(texts, k = 50, max_terms = 1500, min_df = 2) {
  toks <- lapply(texts, .bp_tokens)
  dfreq_all <- table(unlist(lapply(toks, unique)))
  vocab <- names(dfreq_all)[dfreq_all >= min_df]
  if (length(vocab) < 5) return(NULL)
  if (length(vocab) > max_terms)
    vocab <- names(sort(dfreq_all[vocab], decreasing = TRUE))[seq_len(max_terms)]
  vidx <- setNames(seq_along(vocab), vocab)
  n <- length(texts); m <- length(vocab)
  M <- matrix(0, nrow = n, ncol = m)
  for (i in seq_len(n)) {
    tt <- toks[[i]]; tt <- tt[tt %in% vocab]
    if (length(tt)) { tb <- table(tt); M[i, vidx[names(tb)]] <- as.numeric(tb) }
  }
  idf <- log((1 + n) / (1 + colSums(M > 0))) + 1   # TF-IDF weighting
  M <- sweep(M, 2, idf, `*`)
  kk <- min(k, m - 1L, n - 1L)
  if (kk < 2) return(NULL)
  sv <- tryCatch(svd(M, nu = kk, nv = 0), error = function(e) NULL)
  if (is.null(sv)) return(NULL)
  V <- sv$u[, seq_len(kk), drop = FALSE] %*% diag(sv$d[seq_len(kk)], kk, kk)
  nrm <- sqrt(rowSums(V^2)); nrm[nrm == 0] <- 1
  V / nrm
}
```

```{r}
#| eval: false
#| code-summary: "Clearance per post and the per-round bypass measure"
bypass_clearance <- function(df) {
  posts <- df |> dplyr::filter(channel %in% public_unmonitored)
  if (nrow(posts) == 0) return(dplyr::mutate(posts, clearance = numeric(0),
    bypass = numeric(0), best_round = integer(0),
    best_match = character(0), best_channel = character(0)))
  content_all <- as.character(df$content); agent_all <- as.character(df$agent_id)
  round_all <- df$round_idx; chan_all <- as.character(df$channel)

  V <- tryCatch(build_lsa(content_all), error = function(e) NULL)
  tok_all <- if (is.null(V)) lapply(content_all, .bp_tokens) else NULL
  mon_mask  <- chan_all %in% monitored_channels
  post_rows <- which(chan_all %in% public_unmonitored)
  cl <- numeric(length(post_rows)); br <- rep(NA_integer_, length(post_rows))
  bm <- rep(NA_character_, length(post_rows)); bc <- rep(NA_character_, length(post_rows))
  for (ii in seq_along(post_rows)) {
    i <- post_rows[ii]
    # candidates: same agent, monitored channel, this round or earlier
    cand <- which(mon_mask & agent_all == agent_all[i] & round_all <= round_all[i])
    if (length(cand) == 0) next
    sims <- if (!is.null(V)) as.numeric(V[cand, , drop = FALSE] %*% V[i, ])
            else vapply(cand, function(j) .bp_cosine(tok_all[[i]], tok_all[[j]]), numeric(1))
    sims[is.na(sims)] <- 0
    k <- which.max(sims)
    cl[ii] <- max(0, sims[k]); br[ii] <- as.integer(round_all[cand[k]])
    bm[ii] <- content_all[cand[k]]; bc[ii] <- chan_all[cand[k]]
  }
  posts |> dplyr::mutate(clearance = cl, bypass = 1 - cl,
    best_round = br, best_match = bm, best_channel = bc)
}

# Per-round bypass: the average bypass likelihood of that round's posts.
score_bypass <- function(df) {
  b <- bypass_clearance(df)
  if (nrow(b) == 0) return(tibble::tibble(round_idx = integer(0),
                                          posts = integer(0), pct = numeric(0)))
  b |> dplyr::group_by(round_idx) |>
    dplyr::summarise(posts = dplyr::n(), pct = mean(bypass), .groups = "drop")
}
```

## Objective 3, network difference, showing how the structure changed

The network signal treats communication as relationships rather than counts. Directed agent-to-agent links are built from the recipient lists and weighted by message count. For any two rounds the analyst selects, the two link sets are compared so that a link reads as new, dropped or unchanged between them, and node size carries each agent's number of connections. Holding the layout fixed means a change in the wiring is read as a change in colour rather than position.

In the app each round's links are built from two sources, the named recipients of a message and its replies, where a reply is resolved to the author of the message it answers. The reply links matter because many internal messages name no recipient, so without them the baseline would look empty. The links are cached once per dataset so the sliders stay fast. Each round is then checked against the baseline pattern in both directions. A link counts toward the score when it is brand new, never present in any baseline round, or when it is a reliable baseline link, one active in at least half the baseline rounds, that falls silent that round, which is the case of a normally-open connection suddenly dropping. The minimum-messages-per-link control only removes one-off links and is not itself treated as abnormal.

To be exact, the app holds two reference sets from the baseline, every link that appeared in any baseline round and the reliable links active in at least half of them. A round scores the count of its links that are in neither set, the new ones, plus the count of reliable links missing from it, the dropped ones.

```{r}
#| eval: false
#| code-summary: "Agent-to-agent edges from recipients and replies"
build_agent_edges <- function(df) {
  rec <- df |>
    dplyr::filter(!is.na(recipients_csv), recipients_csv != "", recipients_csv != "ALL") |>
    tidyr::separate_rows(recipients_csv, sep = ";") |>
    dplyr::mutate(to = unname(recipient_to_agent[recipients_csv])) |>
    dplyr::filter(!is.na(to)) |>
    dplyr::transmute(from = as.character(agent_id), to = to)
  rep <- tibble::tibble(from = character(0), to = character(0))
  if ("responding_to" %in% names(df)) {
    id2agent <- setNames(as.character(messages_tbl$agent_id),
                         as.character(messages_tbl$message_id))
    rep <- df |>
      dplyr::filter(!is.na(responding_to), responding_to != "") |>
      dplyr::mutate(to = unname(id2agent[as.character(responding_to)])) |>
      dplyr::filter(!is.na(to), to != as.character(agent_id)) |>
      dplyr::transmute(from = as.character(agent_id), to = to)
  }
  dplyr::bind_rows(rec, rep) |> dplyr::count(from, to, name = "weight")
}
```

```{r}
#| eval: false
#| code-summary: "Per-round network score: new plus dropped-reliable links"
score_network <- function(df, split, weight_floor = 1) {
  base_rounds <- sort(unique(df$round_idx[df$round_idx <= split]))
  links_in <- function(r) {
    e <- round_edges(r)
    e <- e[e$weight >= weight_floor, , drop = FALSE]
    if (nrow(e) == 0) character(0) else paste(e$from, e$to)
  }
  all_base <- unlist(lapply(base_rounds, links_in))
  base_any <- unique(all_base)
  freq     <- if (length(all_base) == 0) numeric(0)
              else table(all_base) / max(length(base_rounds), 1)
  reliable <- names(freq)[freq >= 0.5]   # active in half the baseline rounds
  rounds <- sort(unique(df$round_idx))
  dplyr::bind_rows(lapply(rounds, function(r) {
    cur <- links_in(r)
    appeared <- length(setdiff(cur, base_any))   # brand-new connections
    dropped  <- length(setdiff(reliable, cur))   # normally-active links gone quiet
    tibble::tibble(round_idx = r, value = appeared + dropped)
  }))
}
```

## Objective 4, topics, connecting signals to the text

To connect the signals to subject matter, every message is tagged with each topic whose keyword pattern its content matches, and the analyst can also search for their own words. The topic signal counts how often the tracked subject is mentioned each round and flags a spike against the baseline rate.

In the app a chosen topic, several topics or a free-text search are resolved into a single pattern, with a search taking precedence over the dropdown, several words combining as alternatives and an empty choice tracking every topic. Each round's mentions are counted from the message text, and the round's score is how far the count exceeds the spike multiple of the baseline-period average, so only a genuine increase shades the lane. A topic absent from the baseline counts any later appearance in full. The tags and the search also drive the evidence reading, where each message's content sits alongside the agent's private reasoning.

To be exact, the baseline rate is the average mentions per baseline round, the total tracked mentions across the baseline rounds divided by the number of baseline rounds. A later round scores its own mention count minus the spike setting times this average, floored at zero, so with a spike of two a round only shades once its mentions pass twice the baseline rate. A topic that never appeared in the baseline has a rate of zero, so any later mention counts in full.

Tracing this breach took more than one search. No single keyword surfaces the whole story, so the full picture was assembled from several terms such as CivicLoom for the merger subject that first appears at round 9, Section 4.3 for the legal override argument at round 16 and SaltWind for the journalist's exposé. Each term surfaces a different thread, and reading them together reconstructs a sequence that any one search on its own would miss.

```{r}
#| eval: false
#| code-summary: "Resolving the topic pattern and scoring the per-round spike"
resolve_topic_pattern <- function(topics = NULL, search = NULL) {
  if (!is.null(search) && nzchar(trimws(search %||% ""))) {   # search wins
    terms <- trimws(strsplit(search, "[,;]")[[1]]); terms <- terms[nzchar(terms)]
    if (length(terms) > 0)
      return(list(label = paste(terms, collapse = ", "),
                  pattern = paste(tolower(terms), collapse = "|")))
  }
  topics <- topics[topics %in% topic_patterns_default$topic]
  if (length(topics) > 0) {                                   # chosen topics combine with OR
    pats <- topic_patterns_default$pattern[match(topics, topic_patterns_default$topic)]
    return(list(label = paste(topics, collapse = ", "),
                pattern = paste(pats, collapse = "|")))
  }
  list(label = "all topics",                                  # nothing chosen tracks all
       pattern = paste(topic_patterns_default$pattern, collapse = "|"))
}

score_topic <- function(df, split, pattern = NULL, spike = 2) {
  perround <- if (is.null(pattern) && "embargo_keyword_count" %in% names(df)) {
    df |> dplyr::group_by(round_idx) |>
      dplyr::summarise(value = sum(embargo_keyword_count, na.rm = TRUE), .groups = "drop")
  } else {
    pat <- pattern %||% topic_patterns_default$pattern[topic_patterns_default$topic == "Compliance"]
    df |>
      dplyr::mutate(hit = stringr::str_detect(tolower(dplyr::coalesce(content, "")), tolower(pat))) |>
      dplyr::group_by(round_idx) |> dplyr::summarise(value = sum(hit), .groups = "drop")
  }
  rds <- sort(unique(df$round_idx))
  perround <- tibble::tibble(round_idx = rds) |>
    dplyr::left_join(perround, by = "round_idx") |>
    dplyr::mutate(value = tidyr::replace_na(value, 0))
  base_rate <- mean(perround$value[perround$round_idx <= split])
  if (is.na(base_rate)) base_rate <- 0
  perround |> dplyr::mutate(score = pmax(0, value - spike * base_rate))
}
```

## Bringing the signals together

The four per-round scores are laid out as one matrix, one row per signal and one column per round. Each signal's score is normalised to a zero-to-one severity within its own signal so the four can share a single colour scale, and an Overall row averages them. This matrix is the Case Board heatmap, and the same scores drive the deep-dive tabs, so the overview and the detail always agree.

```{r}
#| eval: false
#| code-summary: "Assembling and normalising the abnormality matrix"
build_matrix <- function(df, split, strict, topic_pattern = NULL,
                         weight_floor = 1, spike = 2, bypass_tbl = NULL) {
  rds <- sort(unique(df$round_idx))
  if (is.null(bypass_tbl)) bypass_tbl <- score_bypass(df)
  parts <- dplyr::bind_rows(
    score_activity(df, split, strict) |>
      dplyr::transmute(round_idx, dimension = "Activity", value, score = value),
    bypass_tbl |>
      dplyr::transmute(round_idx, dimension = "Bypass", value = pct * 100, score = pct * 100),
    score_network(df, split, weight_floor) |>
      dplyr::transmute(round_idx, dimension = "Network", value, score = value),
    score_topic(df, split, topic_pattern, spike) |>
      dplyr::transmute(round_idx, dimension = "Topic", value, score)
  )
  tidyr::expand_grid(round_idx = rds,
                     dimension = c("Activity", "Bypass", "Network", "Topic")) |>
    dplyr::left_join(parts, by = c("round_idx", "dimension")) |>
    dplyr::mutate(value = tidyr::replace_na(value, 0), score = tidyr::replace_na(score, 0)) |>
    dplyr::group_by(dimension) |>
    dplyr::mutate(severity = if (max(score) > 0) score / max(score) else 0) |>
    dplyr::ungroup()
}
```

## From the computation to the results

These four signals, all measured against the chosen baseline, are what the application presents. Working through them in the application produces the results reported on the [Findings](../findings.qmd) page, and the design of that application is described in [Interactive Presentation](interactive-presentation.qmd).
