---
title: "Data Preparation"
subtitle: "Methodology · Phase 1 of 3"
author: "Team HeeHaa"
date: last-modified
---

The methodology runs as the three-stage pipeline set out in the [proposal](../proposal.qmd). Raw data is prepared, behavioural signals are computed against a baseline the analyst chooses, and the signals are shown through four interactive sections. This page covers the first stage.

## The three-stage pipeline

The diagram below maps the whole methodology. Raw data is prepared into tidy tables and reference structures, behavioural signals are computed against the chosen baseline, and those signals are presented through the four interactive sections. This page covers the first stage, with the next two covered under [Analytical Computation](analytical-computation.qmd) and [Interactive Presentation](interactive-presentation.qmd).

```{r}
#| label: methodology-pipeline
#| fig-width: 9
#| fig-height: 3.6
#| code-summary: "Pipeline diagram"
pacman::p_load(DiagrammeR)

grViz("
digraph methodology {
  rankdir = LR;
  graph [fontname = 'Helvetica', nodesep = 0.3, ranksep = 0.7];
  node  [fontname = 'Helvetica', shape = box, style = 'filled,rounded',
         fontsize = 11, color = '#2C5282', fillcolor = '#EBF4FF',
         fontcolor = '#1A202C', margin = '0.12,0.08'];
  edge  [color = '#4A5568', arrowsize = 0.7];

  raw [label = 'MC1 nested JSON', fillcolor = '#FEFCBF', color = '#B7791F'];
  app [label = 'Shiny app', fillcolor = '#FEFCBF', color = '#B7791F'];

  subgraph cluster_1 {
    label = '1  Data Preparation';
    fontname = 'Helvetica'; fontsize = 12; fontcolor = '#1A202C';
    style = 'filled,rounded'; fillcolor = '#F7FAFC'; color = '#CBD5E0';
    prep   [label = 'Flatten and clean'];
    tables [label = 'Three tables'];
    derive [label = 'Hierarchy and topics'];
    prep -> tables -> derive;
  }

  subgraph cluster_2 {
    label = '2  Analytical Computation';
    fontname = 'Helvetica'; fontsize = 12; fontcolor = '#1A202C';
    style = 'filled,rounded'; fillcolor = '#F7FAFC'; color = '#CBD5E0';
    base [label = 'Baseline split'];
    abn  [label = 'Abnormality'];
    byp  [label = 'Bypass'];
    net  [label = 'Network difference'];
    base -> abn;
    base -> byp;
    base -> net;
  }

  subgraph cluster_3 {
    label = '3  Interactive Presentation';
    fontname = 'Helvetica'; fontsize = 12; fontcolor = '#1A202C';
    style = 'filled,rounded'; fillcolor = '#F7FAFC'; color = '#CBD5E0';
    secA [label = 'A Activity'];
    secB [label = 'B Bypass'];
    secC [label = 'C Network'];
    secD [label = 'D Evidence'];
  }

  raw    -> prep;
  derive -> base;
  net    -> secA;
  secD   -> app;
}
")
```

## Key context

The scenario is VAST Challenge 2026 Mini-Challenge 1. A property-technology company named TenantThread runs seven AI agents through a public-relations crisis while a merger with CivicLoom sits under embargo. A compliance agent named the Judge watches a subset of channels for violations. The analytical question is whether embargoed content reached the outside world through channels the Judge could not see, and if so, when and by whom.

| Element | Value |
|---|---|
| Rounds in the timeline | 23, from 17 May to 5 June |
| Agents | 7 (Legal, Platform Trust, PR, Social Media, PR Intern, Intern, Judge) |
| Channels | 6, from monitored internal to unmonitored public |
| Messages | 912 |
| Embargo lift | 6:00 PM on 5 June, the final round |

## Source data

The data arrives as a nested JSON file holding 23 timeline rounds. Each round records an event headline, a short narrative and the messages exchanged in that round. The seven agents communicate over six channels of decreasing accountability, and the Judge was assigned to the communications huddle partway through the timeline as the compliance monitor.

## Flattening to three tables

The nested timeline was flattened into three working tables. A message table of 912 rows holds one row per message with its sender, channel, recipients, content and the agent's three private reasoning fields named reacting, rationalizing and deliberating. A round table of 23 rows holds each round's headline and narrative. A network table holds the directed links between agents derived from the recipient lists.

Types were fixed during the read so that every downstream axis orders correctly and every logical flag stays logical. Agent identity and channel were read as factors with explicit level orders, round and word counts as integers, and the embargo and recipient flags as logicals.

```{r}
#| label: load-data
#| eval: false
#| code-summary: "Reading the cleaned message table with fixed column types"

library(tidyverse)

messages_tbl <- readr::read_csv(
  "data/messages_clean.csv",
  col_types = readr::cols(
    round_idx        = readr::col_integer(),
    round_hour_dt    = readr::col_datetime(),
    agent_id         = readr::col_factor(levels = c(
      "legal_agent","quality_agent","pr_agent","social_media_agent",
      "pr_intern_agent","intern_agent","judge_agent")),
    channel          = readr::col_factor(levels = c(
      "comms_huddle","one_on_one_chat","side_huddle",
      "official_post","personal_post","anonymous_post")),
    embargo_keyword_count = readr::col_integer(),
    recipients_is_all     = readr::col_logical(),
    embargo_hit           = readr::col_logical(),
    .default              = readr::col_character()
  )
)
```

## Data dictionary

The message table is the working unit of the analysis. Its main fields are below. Each row is one message.

| Field | Type | Meaning |
|---|---|---|
| `round_idx` | integer | Round number, 1 to 23 |
| `round_hour_dt` | datetime | In-scenario hour the round represents |
| `message_id` | character | Unique identifier for the message |
| `agent_id` | factor | The agent that sent the message |
| `channel` | factor | Channel the message was sent on, from the accountability ladder |
| `message_type` | character | Broadcast or reply |
| `responding_to` | character | The `message_id` this message replies to, if any |
| `recipients_csv` | character | Recipient short names separated by semicolons, or `ALL` for a broadcast |
| `recipients_is_all` | logical | Whether the message went to everyone |
| `content` | character | The visible message text |
| `reacting` | character | The agent's private reaction to the moment |
| `rationalizing` | character | The agent's private justification of its choice |
| `deliberating` | character | The agent's private weighing of options |
| `word_count` | integer | Words in `content` |
| `embargo_keyword_count` | integer | Count of embargo-related keywords found in `content` |
| `embargo_hit` | logical | Whether the message carried embargo content |

The round table holds one row per round with its `event_headline` and `event_narrative`, and the network table holds the directed `from` and `to` agent links with a message-count `weight`. Several analysis fields, such as the calm-period channel share and a deviance flag, were added to the message table during the computation stage and are described under [Analytical Computation](analytical-computation.qmd).

## The channel accountability ladder

The central reference structure is the channel accountability ladder. Every channel was ranked from most to least accountable and marked for whether the Judge can see it. A legitimate announcement should travel down this ladder, starting as internal discussion on a monitored channel and ending as an attributed public post. Content that reaches an unmonitored public channel without that accountable path is the signal the rest of the analysis looks for.

```{r}
#| label: channel-ladder
#| fig-height: 2.8
#| code-summary: "Building and drawing the accountability ladder"
library(tidyverse)

ladder <- tibble(
  channel   = c("Communications huddle","One-on-one chat","Side huddle",
                "Official post","Personal post","Anonymous post"),
  rank      = 1:6,
  layer     = c("Internal","Internal","Internal","Public","Public","Public"),
  monitored = c(TRUE, TRUE, FALSE, FALSE, FALSE, FALSE)
)

ggplot(ladder, aes(x = 7 - rank, y = reorder(channel, 7 - rank), fill = monitored)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = layer), hjust = 1.1, colour = "white", size = 3.2) +
  scale_fill_manual(values = c(`TRUE` = "#2F855A", `FALSE` = "#9B2C2C"),
                    labels = c(`TRUE` = "Monitored by the Judge", `FALSE` = "Unmonitored"),
                    name = NULL) +
  labs(x = "Accountability (longer bar means more visible to the Judge)",
       y = NULL, title = "The six channels, ranked by accountability") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_blank(), panel.grid = element_blank(),
        legend.position = "top")
```

## Other reference structures

Two further structures were prepared. A recipient lookup maps the short recipient names such as legal or pr to the full agent identities such as legal_agent or pr_agent, so that network edges connect to real nodes rather than dropping out. A topic vocabulary gives each of six topics, covering legal, product trust, merger, media, market and compliance language, a keyword pattern so that every message can be tagged by the topics its content matches.

```{r}
#| label: reference-structures
#| eval: false
#| code-summary: "The recipient lookup and the topic vocabulary"

recipient_to_agent <- c(
  legal = "legal_agent", platform_trust = "quality_agent", pr = "pr_agent",
  social_manager = "social_media_agent", pr_intern = "pr_intern_agent",
  intern = "intern_agent", judge = "judge_agent")

topic_patterns <- tibble::tribble(
  ~topic,           ~pattern,
  "Legal",          "\\b(legal|counsel|consent|regulator|attorney|liability)\\b",
  "Product trust",  "\\b(governance|audit|score|data|platform|algorithm)\\b",
  "Merger",         "\\b(merger|civicloom|harborcrest|acquisition|deal)\\b",
  "Media",          "\\b(saltwind|press|journalist|story|sentiment|denial)\\b",
  "Market",         "\\b(market|stock|investor|residentiq|share)\\b",
  "Compliance",     "\\b(embargo|monitoring|judge|compliance|oversight|policy)\\b")
```

## Data overview

With the tables in place, a first look confirms the shape of the data before any signal is computed. The three views below report the spread of messages across agents, across channels, and over the timeline.

```{r}
#| label: overview-setup
#| message: false
#| code-summary: "Reading the cleaned tables for the overview"
messages <- read_csv("data/messages_clean.csv", show_col_types = FALSE)
rounds   <- read_csv("data/rounds_clean.csv",   show_col_types = FALSE)

agent_labels <- c(
  legal_agent = "Legal", quality_agent = "Platform Trust", pr_agent = "PR",
  social_media_agent = "Social Media", pr_intern_agent = "PR Intern",
  intern_agent = "Intern", judge_agent = "Judge")
chan_levels <- c("comms_huddle","one_on_one_chat","side_huddle",
                 "official_post","personal_post","anonymous_post")
```

The seven agents are far from equal contributors. The Judge, as a monitor rather than a participant, speaks rarely, while the Legal and Social Media agents carry the most traffic.

```{r}
#| label: messages-per-agent
#| fig-height: 3.0
messages |>
  count(agent_id) |>
  mutate(agent = recode(agent_id, !!!agent_labels)) |>
  ggplot(aes(x = n, y = reorder(agent, n))) +
  geom_col(fill = "#2C5282") +
  geom_text(aes(label = n), hjust = -0.2, size = 3.5) +
  expand_limits(x = max(table(messages$agent_id)) * 1.1) +
  labs(x = "Messages", y = NULL, title = "Message volume by agent") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank())
```

Most traffic sits on the two monitored internal channels. The unmonitored public channels carry far less volume overall, which is why a small burst of activity on them late in the timeline stands out.

```{r}
#| label: messages-per-channel
#| fig-height: 3.0
messages |>
  mutate(channel = factor(channel, levels = chan_levels)) |>
  count(channel) |>
  ggplot(aes(x = channel, y = n,
             fill = channel %in% c("comms_huddle","one_on_one_chat"))) +
  geom_col() +
  scale_fill_manual(values = c(`TRUE` = "#2F855A", `FALSE` = "#9B2C2C"),
                    labels = c(`TRUE` = "Monitored", `FALSE` = "Unmonitored"),
                    name = NULL) +
  labs(x = NULL, y = "Messages", title = "Message volume by channel") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1), legend.position = "top")
```

Messages per round separate the two regimes in the data. The pre-crisis rounds carry a steady, moderate load. From round 14 onward, the crisis day on 5 June, the rounds advance hour by hour and volume climbs as the story breaks publicly.

```{r}
#| label: activity-over-time
#| fig-height: 3.0
rounds |>
  ggplot(aes(x = round_idx, y = n_messages)) +
  geom_col(fill = "#4A5568", width = 0.8) +
  geom_vline(xintercept = 13.5, linetype = "dashed", colour = "#C05621") +
  annotate("text", x = 13.5, y = max(rounds$n_messages),
           label = " crisis day begins", hjust = 0, size = 3.2, colour = "#C05621") +
  scale_x_continuous(breaks = seq(1, 23, 2)) +
  labs(x = "Round", y = "Messages", title = "Messages per round across the timeline") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

Tagging messages with the merger keyword pattern shows the topic vocabulary in action. The merger is discussed on the monitored internal channels through the build-up, and the dark red bars, the anonymous channel, appear only in the closing rounds.

```{r}
#| label: merger-topic
#| fig-height: 3.0
channel_palette <- c(
  comms_huddle = "#90CDF4", one_on_one_chat = "#F6AD55", side_huddle = "#FBD38D",
  official_post = "#68D391", personal_post = "#FC8181", anonymous_post = "#742A2A")
merger_pat <- "\\b(merger|civicloom|harborcrest|acquisition|deal)\\b"

messages |>
  mutate(text = tolower(coalesce(content, ""))) |>
  filter(str_detect(text, merger_pat)) |>
  mutate(channel = factor(channel, levels = chan_levels)) |>
  count(round_idx, channel, name = "mentions") |>
  ggplot(aes(round_idx, mentions, fill = channel)) +
  geom_col(width = 0.8) +
  scale_fill_manual(values = channel_palette, drop = FALSE, name = NULL) +
  scale_x_continuous(breaks = seq(1, 23, 2)) +
  labs(x = "Round", y = "Mentions", title = "Merger-topic mentions by round and channel") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom")
```

The preparation stage produces the clean message, round and network tables together with the accountability ladder, the recipient lookup and the topic vocabulary. These feed the signal computation in [Analytical Computation](analytical-computation.qmd).
